Electron and ion acceleration by relativistic shocks: particle-in-cell model simulations

### 2017-11-17 papers/Physica2017/Romansky_Physica2017.tex
paper

\begin{document}
\title{Particle-in-cell simulation of ion and electron acceleration in trans-relativistic shocks.}

\author{V I Romansky$^{1,2}$, A M Bykov$^1$, S M Osipov$^1$ and P E Gladilin$^1$}

\address{$^1$ Ioffe Institute, 26 Politekhnicheskaya st., St. Petersburg 194021, Russia}
\address{$^2$ 
Sternberg Astronomical Institute, Moscow State University
 Universitetsky pr., 13, Moscow 119234, Russia}

@@

\begin{abstract}
The problem of Cosmic Ray (CR) origin and acceleration is a widely discussed problem of
modern astrophysics. It is considered that most likely CR sources are supernova remnant
shock waves. One of the methods to study particle injection and acceleration by shock
waves is a particle-in-cell computer simulation. In this paper we present results of shock
wave simulations based on our self-developed implicit particle-in-cell code. 
\end{abstract}
\input{intro.tex}
\input{model.tex}
\input{test.tex}
\input{shock.tex}
\input{conclusions.tex}
\input{biblio.tex}
\end{document}

>>>>

\begin{document}
\title{Electron and ion acceleration by relativistic shocks: particle-in-cell model simulations}

\author{V I Romansky$^{1,2}$, A M Bykov$^{1,3}$ and S M Osipov$^1$}

\address{$^1$ Ioffe Institute, 26 Politekhnicheskaya st., St. Petersburg 194021, Russia}
\address{$^2$
	Sternberg Astronomical Institute, Moscow State University
	Universitetsky pr., 13, Moscow 119234, Russia}
\address{$^3$ Peter the Great St.~Petersburg Polytechnic University, 29 Politekhnicheskaya st., St. Petersburg 195251, Russia}

@@

\begin{abstract}
	The problem of particle acceleration by the collisionless astrophysical shocks is of fundamental importance for the cosmic ray origin problem and the high energy astrophysics. The fast magnetohydrodynamical shocks produced by relativistic jets are considered as the very likely sites of the cosmic ray acceleration.
	While cosmic ray acceleration by both non-relativistic and ultra relativistic shocks was studied in some details the shocks of a moderate Lorentz factors has attracted much less attention so far, while they are important for modelling the gamma-ray burst afterglows and a number of other interesting objects. In this paper we present the results of simulations of relativistic shock waves in the electron-proton plasmas obtained with an implicit particle-in-cell code for the initial study the particle injection efficiencies.
\end{abstract}
\input{intro.tex}
\input{shock.tex}
\input{conclusions.tex}
\input{biblio.tex}
\end{document}

### 2017-11-17 papers/Physica2017/conclusions.tex
paper

\section{Conclusions}
In this paper we studied trans-relativistic shock wave using our implicit particle-in-cell code. Simulations show, that particles injection in acceleration process highly depends on angle between flow bulk velocity and magnetic field. Also injection of protons is more effective than electrons for angles less then critical, but there is regime of more effective electrons acceleration when angle is close to critical. Further research is necessary in order to study this phenomena in detail.
\ack
V I Romansky acknowledges support from RSF grant 16-12-10519.
The results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.spbstu.ru).

>>>>

\section{Conclusions}
In this paper we presented the first results of the PIC simulations of particle acceleration by the moderately relativistic shock waves. Simulations show, that the injection of particles into the acceleration process depends strongly on the angle between the flow bulk velocity and the magnetic field. The injection efficiency as a function of the inclination angle is studied. The injection of protons is more efficient than that of electrons for the shock inclination angles below the critical angle where the transition to the superluminal shock regime occurs. On the other hand the electron acceleration is more efficient in the shocks with the inclination angles which are close to the critical angle.
\ack
V I Romansky acknowledges support from RSF grant 16-12-10519.
The results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://www.scc.spbstu.ru).

### 2017-11-17 papers/Physica2017/intro.tex
paper

\section{Introduction}
One of the possible sources of high-energy cosmic rays are shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}. The particle acceleration on the shock front is sophisticated and nonlinear problem, because cosmic rays have strongly influence on the shock wave. The pressure of cosmic rays is comparable with pressure of background plasma and significatly modifies the front (see for example \cite{Bykov2014}). Also there are a instabilities in anisotropic plasma, which lead to strong field amplification and the observations of the synchrotron emission from supernova remnants (SNRs) show, that the magnetic field there is about 100 times stronger then the interstellar field (see for example \cite{Berezhko2003},\cite{Uchiyama2007}). One of the most powerful methods to explore processes in collisionless plasma is the Particle-in-Cell simulation.

>>>>

\section{Introduction}
One of the possible sources of the high-energy cosmic rays are the shock waves in astrophysical collisionless plasma due to the diffusive shock acceleration (DSA) mechanism, see \cite{Bell1978}, \cite{Blandford1978}. The particle acceleration on the shock front is a sophisticated and nonlinear problem, because the cosmic rays have a strong influence on the shock wave. The pressure of cosmic rays is comparable with the pressure of background plasma and significantly modifies the front (see, e.g., \cite{Bykov2014}). 
Also the instabilities in the anisotropic plasma can yield a strong field amplification, and according to the observations of synchrotron emission from supernova remnants (SNRs), the magnetic field there is about 100 times stronger than the interstellar field (see, e.g., \cite{Berezhko2003},\cite{Uchiyama2007}). 
One of the most powerful methods to explore the processes in collisionless plasma is the Particle-in-Cell simulation.

### 2017-11-17 papers/Physica2017/shock.tex
paper

\section{Shock wave simulation}
We studied the evolution of trans-relativistic shocks and spectum of accelerated particles. We chose relastivistic shocks with low lorenz-factor because the maximum energy of produced cosmic rays increases with shock wave lorenz-factor, but the efficiency of acceleration decreases, because it is more difficult for particle to cross fast moving front many times \cite{Ellison2013}. So we assume that intermediate case of trans-relativistic shocks provides the most effective acceleration.

We developed the implicit particle-in-cell (PIC) code, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
Our code is fully three dimensional and parallelized with MPI technology, which is adapted for distributed computing and can be executed on a wide class of computers.

In simulation setup the homogeneous plasma, which flows in through the right boundary
and collides with the reflecting superconducting wall on the left boundary.

 Simulations are one-dimensional and have following parameters: initial flow lorenz-factor $\gamma = 1.5$, number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , temperature $5\cdot10^8 \rm{K}$, magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 4\cdot10^{12} \rm{cm}$, the number of cells $N=4\cdot10^4$. Electron mass is reduced to $m_e = \frac{m_p}{20}$. The full time of simulation is $T = 5000 {\omega_p}^{-1}$. All values are measured in downstream frame. $\theta$ is angle between flow velocity and magnetic field. We present results of particle spectrum in several simulations with different $\theta$. 
\begin{figure}[h!]
	\centering
	\includegraphics[width=0.8\textwidth]{fig/protons.eps} 
	\caption{Distribution of protons.}
	\label{protons}
\end{figure}
\begin{figure}[h!]
	\centering
	\includegraphics[width=0.8\textwidth]{fig/electrons.eps} 
	\caption{Distribution of electrons.}
	\label{electrons}
\end{figure}

One can see that spectrum strongly depends of angle $\theta$. If $\theta$ is less then critical value, defined as $c\cdot cos(\theta_{crit})=v_{shock}$, measured in shock rest frame, shock is subluminal and particles can escape from front and than cross it enough of times to e accelerated.

>>>>

\section{Shock wave simulation}
We studied the evolution of moderately relativistic shocks and spectrum of accelerated particles. We chose relativistic shocks with a low Lorentz factor because the maximum energy of the produced cosmic rays increases with the shock wave Lorentz factor, while the efficiency of acceleration decreases, 
because it is more difficult for particle to cross fast moving front many times 
%
\cite{Ellison2013}. So we assume that the intermediate case of moderately relativistic shocks provides the most efficient acceleration.

We developed the implicit particle-in-cell (PIC) code, presented in our previous paper \cite{Romansky2016}, based on the scheme suggested by Lapenta et al.~\cite{Lapenta2006} and improved for the relativistic case by Noguchi et al.\cite{Noguchi2007}.
Our code is fully three-dimensional and parallelized with MPI technology, which is adapted for distributed computing and can be executed on a wide class of computers. We use 
%cleaning divergence of electric field 
the electric field divergence correction
after every several time steps and 
%Fourier filtering of shortwave harmonics 
the shortwave harmonics Fourier filtering
to suppress the numerical Cherenkov instability and improve the energy conservation.

In the modeling setup the homogeneous plasma flows in the simulation box through the right boundary
and collides with the reflecting superconducting wall on the left boundary, launching the shock wave. It is a common way to initialize a shock wave, because the initialization using the Rankine–Hugoniot conditions does not take into account the microscopic distribution, and the shock wave, created in such way, will probably fall apart into several discontinuities.

\begin{figure}[h!]
	\centering
	\includegraphics[width=1.0\textwidth]{fig/protons.eps} 
	\caption{Distribution of protons in the relativistic shock wave with Lorentz factor 1.5 with different inclination angle.}
	\label{protons}
\end{figure}

The simulations are one-dimensional and have the following parameters: the initial flow Lorentz factor $\gamma = 1.5$, the number densities $n_e = 10^{-4} \rm{cm}^{-3}$, $n_p = 10^{-4} \rm{cm}^{-3}$ , the temperature $5\cdot10^8 \rm{K}$, the magnetic field $B = 10^{-4} \rm{G}$, the full size of the box $L = 2\cdot10^{12} \rm{cm}$, the number of cells $N=2\cdot10^4$. The electron mass is reduced to $m_e = \frac{m_p}{20}$. The full time of simulation is $T = 5000 {\omega_p}^{-1}$.This values gives the dimensionless parameter magnetization $\sigma = \frac{B^2}{4\pi\gamma (n_p m_p + n_e m_e) c^2} = 0.003$. Inclination angle $\theta$ is the angle between the flow velocity and the magnetic field. We present the results for the particle spectrum in several simulations with different $\theta$. The stable shock waves formed in all simulations and the total energy deviation is less than $3\%$, and it proves that our code provides correct model of the shock waves.

\begin{figure}[h!]
	\centering
	\includegraphics[width=1.0\textwidth]{fig/electrons.eps} 
	\caption{Distribution of electrons in the relativistic shock wave with Lorentz factor 1.5 with different inclination angle.}
	\label{electrons}
\end{figure}

One can see in Figures \ref{protons} and \ref{electrons} the particle spectrum for different angles $\theta$. The spectra consist of three parts - the narrow peak of cold but fast moving initial flow, the wide peak of hot downstream and the non-thermal accelerated component. Figures show that the spectrum of accelerated particles strongly depends on angle $\theta$. If $\theta$ is less than critical value, defined by the equation $c\cdot \cos(\theta_{crit})=v_{shock}$, where all values are measured in the upstream rest frame, particles can escape from the front and cross it several times to gain more energy (see, e.g., \cite{Pelletier2017}). It is difficult to evaluate the critical angle a priori, because the compression relation of the shock wave which will be created in the simulation is unknown. However the estimation for case of a strong wave (the compression relation equals $4$) provides $\theta_{crit}=45^{\circ}$ in the downstream frame, which is consistent with the results of our simulations.

Also, one can see that for the angles less then the critical one, the particle spectra are higher and longer for larger $\theta$. The spectrum of non-thermal component is about 10 times higher for protons than for electrons at the same energy. It is consistent with other Particle-in-cell simulations of the shock waves {\cite{Sironi2011}}.

The results show some interesting effects for angles close to the critical value (black and yellow lines)- electrons are still accelerating while protons are not. It can be explained by electrons acceleration in the short-wavelength turbulence of the shock wave precursor. Due to the small gyroradius of electrons in initial flow, short-wavelength turbulence scatter them more efficiently than protons. However, further research is necessary in order to study this phenomena in detail.
